Species motif participation provides unique information about species risk of extinction

### 2022-05-25 manuscript/reorder_main_text.tex
Updates from Overleaf

    %\subsection*{2 Global and local network properties are known to affect stability}
    How population decline or extinction of one species leads to loss of other species via direct and indirect effects (i.e., secondary extinctions) is a vibrant area of research \citep{Santos2021,curtsdotter2011robustness, dunne2009cascading, Eklof2006}.
    Several global (whole-network) as well as local (single-species) properties are known to influence how species respond to disturbances. 
%network functioning \citep{Petchey2002}, stability \citep{Allesina2012} and robustness to disturbances \citep{Dunne2002, Eklof2006}. 
    Two well-known, influential global properties are network size, the number of species present in the network, and network connectance, the number of realized feeding interactions.
    Based on simulations of random networks, large networks were initially thought to be less stable~\citep{May1972}.
    The presence of large, stable, empirical networks has spurred research into how other structural characteristics, including connectance, may allow large numbers of species to persist~\citep{Dunne2002}.
    Highly-connected networks may be more resistant to secondary extinctions \citep{Dunne2002, Eklof2006} because consumers, on average, have access to a larger number of prey species. However, in a well-connected network a disturbance may spread more easily and cause more secondary extinctions due to the presence of many possible pathways.
    The net effect of connectance on extinction therefore likely depends on the severity of disturbance and the ability of consumers to subsist on a reduced set of prey.

    Local properties known to affect species' response to disturbances include trophic level (a measure of the path length between a consumer and a basal resource) and in-degree (number of prey items of a predator). 
    Species with high trophic levels (long paths to basal resources) and low in-degree (few prey) are generally at higher risks of secondary extinction \citep{binzer2011susceptibility, Eklof2006}, including secondary extinction due to reduced plant populations. 
    However, in-degree and trophic level are both relatively coarse descriptions of a species' role in its food web~\citep{Cirtwill2018FoodWebs}. 
    As well as the properties of the consumer itself, a consumer's risk of secondary extinction after a disturbance to plants depends on the extinction risk of its prey.
    For example, a generalist consuming many specialist (and therefore high-risk) prey may, for example, be at greater risk of extinction than a specialist which consumes a highly generalist resource that is very unlikely to go extinct. 
    %For example, a consumer with low trophic level may be at high risk of extinction if it consumes prey which are themselves very likely to be affected by disturbance,and a specialist may not necessarily have a high risk of extinction if it consumes a basal resource which is unlikely to be affected by disturbance.

@@

    While the potential relationships between motif participation and species-level extinction risk have not yet been tested to our knowledge, it has been shown that the frequencies of different motifs within a whole network are associated with community stability \citep{prill2005dynamic, bascompte2005simple} and with changes to the composition of plant communities \citep{giling2019plant}.
    While these studies uncovered the potential influence of network-level motif frequencies on network stability, a disturbance may likely alter the network and the interactions between species in different ways. 
    This can affect the influence of motif frequencies on network stability, as it has been demonstrated that an unstable motif may appear stabilizing dependent on interactions with other species \citep{Kratina2012}.

%\subsection*{Here we test...}
    Therefore, we are interested in whether a consumer's extinction risk depends on 1) the motifs in which the species participates and 2) the motif profile of the full food web, in scenarios with various levels of disturbance targeting the basal level.
    In addition, we want to understand how the information provided by motifs is related to other global and local food web properties in order to better synthesize information provided by different scales of network structure.
    We model extinction probabilities using Bayesian network modeling \citep{Eklof2013, Haussler2020}
    Of the 13 three-species motifs that can occur in empirical food webs, four motifs are over-represented in empirical networks and are thought to contribute to stability; the three-species chain, omnivory, apparent competition, and direct competition \citep{Stouffer2007, Borrelli2015a, giling2019plant}.
    Therefore, we focus on these four particularly important motifs.

>>>>

    %\subsection*{1 Why are bottom-up effects important for understanding stability?}
     The plant community is the foundation upon which the myriad species in terrestrial food webs depend for their survival. Disturbances in the plant community have been shown to affect important ecosystem properties such as primary \citep{Hector1999} and secondary production \citep{borer2012plant}, soil respiration and carbon cycling \citep{chen2019plant} and consumer diversity \citep{scherber2010bottom, Baiser2016}.
     As such, the structure of the plant community affects stability across all levels of ecosystem organization, which is crucial for sustaining a healthy energy flow in the ecosystem as a whole~\citep{proulx2010diversity, scherber2010bottom, Rosenblatt2016}. 
     For example, the loss of diversity at the basal level typically causes declines in both abundance and richness of all types of consumers: herbivores, predators, parasitoids, etc. \citep{scherber2010bottom}. Secondary extinctions and plants...

    %\subsection*{2 Global and local network properties are known to affect stability}
    How disturbances, such as population decline or extinction, lead to loss of other species via direct and indirect effects (i.e., secondary extinctions) is a vibrant area of research \citep{Santos2021,curtsdotter2011robustness, dunne2009cascading, Eklof2006}.
    Several global (whole-network) as well as local (single-species) properties are known to influence how species respond to disturbances. Two well-known, influential global properties are network size, the number of species present in the network, and network connectance, the number of realized feeding interactions.
    Based on simulations of random networks, large networks were initially thought to be less stable~\citep{May1972}.
    The presence of large, stable, empirical networks has spurred research into how other structural characteristics, including connectance, may allow large numbers of species to persist~\citep{Dunne2002}.
    Highly-connected networks may be more resistant to secondary extinctions \citep{Dunne2002, Eklof2006} because consumers, on average, have access to a larger number of prey species. However, in a well-connected network a disturbance may spread more easily and cause more secondary extinctions due to the presence of many possible pathways \citep{}.
    The net effect of connectance on extinction therefore likely depends on the severity of disturbance and the ability of consumers to subsist on a reduced set of prey.

    Local properties known to affect species' response to disturbances include trophic level (a measure of the path length between a consumer and a basal resource) and in-degree (number of prey items of a predator). 
    Species with high trophic levels (long paths to basal resources) and low in-degree (few prey) are generally at higher risks of secondary extinction \citep{binzer2011susceptibility, Eklof2006}, including secondary extinction due to reduced plant populations.

    However, in-degree and trophic level are both relatively coarse descriptions of a species' role in its food web~\citep{Cirtwill2018FoodWebs}. 
    As well as the properties of the consumer itself, a consumer's risk of secondary extinction after a disturbance to plants depends on the extinction risk of its prey.
    For example, a generalist consuming many specialist (and therefore high-risk) prey may, for example, be at greater risk of extinction than a specialist which consumes a highly generalist resource that is very unlikely to go extinct.

@@

    While the potential relationships between motif participation and species-level extinction risk have not yet been tested to our knowledge, it has been shown that the frequencies of different motifs within a whole network are associated with community stability \citep{prill2005dynamic, bascompte2005simple} and with changes to the composition of plant communities \citep{giling2019plant}. This indicate that motifs may provide additional information on network resilience to disturbances.

    The severity of disturbances impacting structure forming species, i.e., plant species, is important for how the food web respond \citep{}. Naturally, the magnitude of the effect will increase with severity of the disturbance \citep{}, but additionally disturbances impacting structure forming species, i.e., plant species, may fundamentally alter food web structure \citep{byrnes2011climate}. This has been shown in several studies where the magnitude of the disturbances alter the competition between species in the benthic community \citep{detmer2021variation} and change the percent coverage of different taxonomic groups \citep{carnell2020more} in kelp forests.

    While these studies uncovered the potential influence of network-level motif frequencies on network stability, a disturbance may likely alter the network and the interactions between species in different ways. 
    This can affect the influence of motif frequencies on network stability, as it has been demonstrated that a previously unstable motif may appear stabilizing dependent on interactions with other species \citep{Kratina2012}.

%\subsection*{Here we test...}
   Here, we are interested in whether a consumer's extinction risk of going secondary extinct depends on 1) the motifs in which the species participates, 2) the motif profile of the full food web and 3) the severity of the disturbance. %We address this in scenarios with various levels of disturbance targeting the basal level.
    In addition, we want to understand how the information provided by motifs is related to other global and local food web properties in order to better synthesize information provided by different scales of network structure.
    We model secondary extinction probabilities using Bayesian networks \citep{Eklof2013, Haussler2020}. 
    We focus our analyse on the four network motifs that are over-represented in empirical networks and additionally have been shown to contribute to network stability; the three-species chain, omnivory, apparent competition, and direct competition \citep{Stouffer2007, Borrelli2015a, giling2019plant}.
